Article Title

### 2013-10-25 plos.tex
Revised results to be more concise. Removed details still need to be put back into the methods. Fixed my name & affiliations

\DeclareMathOperator{\Tr}{tr}
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\mcond}{\,\middle\vert\,}
\newcommand{\cond}{\,\vert\,}
\newcommand{\loss}[1]{\mathcal L\left(#1\right)} 
\newcommand{\T}{\intercal}
\newcommand{\E}[2][]{\mathbb E_{#1}\left[ #2\right]}    % expected value
\newcommand{\TODO}[1]{\emph{\small\color{blue}$\langle\langle$#1$\rangle\rangle$}}
\newcommand*\dif{\mathop{}\,d}
\DeclareMathOperator{\rank}{rank}

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
Improved estimates of neural correlations suggest detailed interactions in visual cortex
}
% Insert Author names, affiliations and corresponding author email.
\\
Dimitri Yatsenko$^{1,\ast}$, 
Kre\v{s}imir Josi\'{c}$^{2}$,
Alexander Ecker$^{3,1}$,
Emmanouil Foudarakis$^{1}$,
R.~James Cotton$^{1}$,
Andreas S.~Tolias$^{1}$
\\
\bf{1} Department of Neuroscience, Baylor College of Medicine, Houston, TX, USA
\\
\bf{2} Department of Mathematics and Department of Biology and Biochemistry, University of Houston, Houston, TX, USA
\\
\bf{3} Max Planck Institute of Biological Cybernetics, T\"ubingen, Germany
\\
$\ast$ E-mail: yatsenko@cns.bcm.edu
\end{flushleft}

@@

% Results and Discussion can be combined.
\section*{Results}
\input{sections/Acquisition.tex}
\input{sections/Theory.tex}
\input{sections/Regularization.tex}
\input{sections/Estimators.tex}
\input{sections/Simulation.tex}
\input{sections/Findings.tex}
\input{sections/Structure.tex}

>>>>

\newcommand{\Kcomment}[1]{{\color{blue}{[KJ: #1]}}}
\newcommand{\Acomment}[1]{{\color{red}{[AE: #1]}}}

\DeclareMathOperator{\Tr}{tr}
\newcommand{\mcond}{\,\middle\vert\,}
\newcommand{\cond}{\,\vert\,}
\newcommand{\loss}[1]{\mathcal L\left(#1\right)} 
\newcommand{\T}{{\rm T}}
\newcommand{\E}[2][]{\mathbb E_{#1}\left[ #2\right]}    % expected value
\newcommand{\TODO}[1]{\emph{\small\color{blue}$\langle\langle$#1$\rangle\rangle$}}
\newcommand*\dif{\mathop{}\,d}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator{\rank}{rank}

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
Improved estimates of neural correlations suggest detailed interactions in visual cortex
}
% Insert Author names, affiliations and corresponding author email.
\\
Dimitri Yatsenko,$^{1,\ast}$, 
Kre\v{s}imir Josi\'{c}$^{2}$,
Alexander S.~Ecker$^{1,3,4}$,
Emmanouil Froudarakis$^{1}$,
R.~James Cotton$^{1}$,
Andreas S.~Tolias$^{1,5}$
\\
\bf{1} Department of Neuroscience, Baylor College of Medicine, Houston, TX, USA
\\
\bf{2} Department of Mathematics and Department of Biology and Biochemistry, University of Houston, Houston, TX, USA
\\
\bf{3}  Werner Reichardt Center for Integrative Neuroscience and Institute for Theoretical Physics, University of T\"ubingen, Germany
\\
\bf{4} Bernstein Center for Computational Neuroscience, T\"ubingen, Germany
\\
\bf{5} Department of Computational and Applied Mathematics, Rice University, Houston, TX, USA

$\ast$ E-mail: yatsenko@cns.bcm.edu
\end{flushleft}

@@

% Results and Discussion can be combined.
\section*{Results}

\input{sections/Results.tex}

%\input{sections/Acquisition.tex}
%\input{sections/Theory.tex}
%\input{sections/Regularization.tex}
%\input{sections/Estimators.tex}
%\input{sections/Simulation.tex}
%\input{sections/Findings.tex}
%\input{sections/Structure.tex}

### 2013-10-25 sections/Results.tex
Revised results to be more concise. Removed details still need to be put back into the methods. Fixed my name & affiliations

\subsection*{Data acquisition}
We recorded the calcium activity  of dense populations of neurons in the supragranular layers in primary visual cortex of anesthetized mice using fast random-access 3D scanning two-photon microscopy \cite{Stosiek:2003,Reddy:2005}.  Numerious repetitions of full-field drifting gratings (Fig. 1A and 1B) were presented to the eye contralateral to the imaged site. This technique allowed recording from a large number (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4. Somatic calcium signals were deconvolved using  sparse nonnegative deconvolution \cite{Vogelstein:2010} (Fig.\;1C and 1D).  The average stimulus response was subtracted to remove stimulus covariance and the sample noise covariance matrix was computed (Fig.\;1E).

\input{sections/Figure1.tex}

\subsection*{The sample covariance matrix}
We aim to estimate the true covariance matrix $\Sigma = \E{(x-\mu)(x-\mu)^\T}$ of the instantaneous activity vector $x$ of a population of $p$ neurons. Here $\E{\cdot}$ denotes expectation \TODO{for the true data generating process} and $x$ is the $p\times 1$ vector of real-valued instantaneous firing rates discretized into bins of duration $\Delta t$ and $\mu = \E{x}$.

For more rigorous notation, definitions, and derivations, see Appendix.

The usual estimator of $\Sigma$ is the sample covariance matrix
\begin{equation}
\hat \Sigma_0 = \frac 1 \nu \sum\limits_{t=1}^n (x(t)-\mu)(x(t)-\mu)^\T 
\end{equation}
where $x(t),\;t=1,\ldots,n$ are sequential observations of population activity inferred from calcium signals; $\nu$ is the number of degrees of freedom. For independent observations $\nu=n-1$ because estimation of the mean $\mu$ accounts for one degree of freedom. When observations are correlated, as is the case with calcium signals, $\nu < n-1$ and may be estimated from the signal.

The sample covariance matrix is constructed to be unbiased, such that $\E{\hat\Sigma_0} = \Sigma$.

\subsection*{Evaluation of covariance matrix estimates}
The quality of a covariance matrix estimate $\hat\Sigma$ is measured by a real-valued \emph{loss function} $\loss{\hat\Sigma,\Sigma}$.  The loss function quantifies the deviation of $\hat\Sigma$ from $\Sigma$ and attains its minimum  when $\hat\Sigma = \Sigma$.

For the purposes of this study, we adopted the \emph{negative normal log-likelihood loss} function:
\begin{equation}\label{eq:loss}
\loss{\hat\Sigma,\Sigma} = \frac 1 p\left[\ln \det \hat \Sigma + \Tr(\hat \Sigma^{-1}\Sigma)\right]
\end{equation}
This choice is motivated by mathematical convenience. Other popular choices for the loss function are the Frobenius norm of the difference $\hat\Sigma-\Sigma$, Stein's entropy loss, and quadratic loss \cite{James:1961,Ledoit:2004,Schafer:2005,Fan:2008}.  We assert that the main conclusions of our study are unlikely to change drastically other well behaved loss functions.

The aim of our project is produce covariance matrix estimates that minimize the expected loss.  The expected loss of an estimator is known as its \emph{risk}: 
\begin{equation}
r = \E{\loss{\hat\Sigma, \Sigma}}
\end{equation}

In practice, the true value $\Sigma$ is not accessible and estimators' risks must be estimated from the data.  This may be accomplished through validation: 
Let $\hat\Sigma_0^\prime$ denote a sample covariance matrix measured from an independent sample that was not used included in the computation of $\hat\Sigma$. Then \emph{empirical loss} is 
\begin{equation}
\hat \ell = \loss{\hat\Sigma,\hat\Sigma_0^\prime}
\end{equation}
 and its expected value or \emph{empirical risk} is
\begin{equation}\label{eq:empiricalRisk}
\hat r = \E{\hat\ell} = \mathbb E_{\hat\Sigma} \left[ \mathbb E_{\hat\Sigma_0^\prime} \left[ {\loss{\hat\Sigma,\hat\Sigma_0^\prime}} \right] \right]
\end{equation}
Because the chosen loss function is linear in its second argument in the sense that
\begin{equation}
\loss{\hat\Sigma,X_1} + \loss{\hat\Sigma,X_2} \equiv \loss{\hat\Sigma,X_1+X_2}
\end{equation}
the expection on the second argument in Eq.~\ref{eq:empiricalRisk} may be taken inside the loss function:
\begin{equation}
\hat r = \E{\loss{\hat\Sigma,\E{\hat\Sigma_0^\prime}}}  = \E{\loss{\hat\Sigma,\Sigma}} = r
\end{equation}
Thus the empirical loss $\loss{\hat \Sigma,\hat \Sigma_0^\prime}$ serves as an unbiased estimate of risk $r$.

Because the loss function is equivalent to the negative normal log likelihood, the above derivation led us to the familiar criterion that the optimal covariance matrix estimator is one that consistently maximizes the normal log likelihood of the validation dataset.

\subsection*{Regularization}
Under many loss functions\footnote{
The strict equality in Eq.~\ref{eq:bias-variance} does not hold under the loss function in Eq.~\ref{eq:loss}. However, the equality does hold for its close cousin, Stein's \emph{entropy loss},  which only differs by the order of its arguments and a constant offset: $\mathcal L_s(\hat\Sigma,\Sigma) \equiv \loss{\Sigma,\hat\Sigma} - \loss{\Sigma,\Sigma}$. This defficiency presents no difficulty because we minimize the risk directly, without assessing the two error components individually. The bias-variance decomposition is presented here to motivate the use of regularization.}, 
the estimator's risk can be decomposed as the sum
\begin{equation}\label{eq:bias-variance}
r = b + \varepsilon
\end{equation}
of \emph{approximation error} (``bias'' or systematic error)
\begin{equation}
b = \loss{\bar\Sigma,\Sigma}
\end{equation}
and \emph{estimation error} (``variance'') 
\begin{equation}
\varepsilon = \E{\loss{\hat\Sigma, \bar\Sigma}}
\end{equation}
where $\bar\Sigma = \E{\hat\Sigma}$ is the expected value of the estimate.

The unbiased estimator $\hat\Sigma_0$ makes $\bar\Sigma=\Sigma$ and thereby minimizes the approximation error, but may be excessively susceptible to sample noise and result in high estimation error.

The estimator risk can be reduced by \emph{regularization}. Regularization is the deliberate biasing (\emph{``shrinkage''}) of the estimate toward a low-dimensional, less variable \emph{target estimate} \cite{Bickel:2006,Ledoit:2004}. A regularized estimator solves the bias-variance tradeoff to produce a biased but less variable estimates aiming to minimize the estimator's risk.  Various regularization schemes focus on the dimensionality reduction part \TODO{rephrase} by selecting the optimal target estimate from a family of estimates with reduced dimensionality \cite{findit}.  Other estimators only shrink the sample covariance matrix toward a single target estimator \cite{Schafer:2005}. Yet other regularizers effectively combine shrinkage and dimensionality reduction \cite{findit}.

\input{sections/Estimators.tex}

\subsection*{In simulation, estimator selection `recognizes' the covariance structure}
\input{sections/Figure3.tex}

\subsection*{In visual cortex, Estimator D dominates}
\input{sections/Figure4.tex}
This is the full matrix of all comparisons of the estimators.  The plots represents the histograms of the differences of the loss function between the pair of estimators for all 31 sites.  When the differences are positive, the first estimate in the title outperforms the second estimate.

Of particular interest is the first row, which shows that all regularized estimates  outperformed the sample covariance estimate.  Even more informative is the last column, which shows that estimate D (sparse+lowrank) significantly outperformed all other estimates.

>>>>

\subsection*{Covariance estimation}

Our goal is to estimate the covariance matrix, $\Sigma$, of a population of $n$ neurons. For small populations or in the limit of infinite data this can be done in a straightforward manner by using the sample covariance estimator
\begin{equation}
\hat \Sigma_0 = \frac 1 \nu \sum\limits_{t=1}^n (x(t)-\mu)(x(t)-\mu)^\T, 
\end{equation}
where $x(t)$ are sequential observations of population activity and $\nu$ is the number of degrees of freedom ($\nu=n-1$ if observations are independent).

However, for large population sizes and limited data this approach becomes ineffective at capturing the true structure of the covariance matrix (REFS). The number of parameters to be estimated grows quadratically with the population size. Thus, to obtain a "good" estimate of the true covariance, we need to rely on regularization -- the deliberate biasing of the estimate toward a low-dimensional, less variable target estimate \cite{Bickel:2006,Ledoit:2004}. Various such targets exist, each expressing a different idea about what a parsimonious model would look like.

We considered four different estimators, based on a few simple concepts for regularizing the covariance: independence, low-rank structure and sparse interactions. We optimized all four estimators with respect to Gaussian log-likelihood (see Discussion for a more in-depth justification of this loss function) and used cross-validation to determine hyper-parameters.

The first and probably simplest regularized estimator is shrinkage towards a diagonal matrix
\begin{equation}
\hat \Sigma_{\rm diag} = (1-\lambda) \hat \Sigma_0 + \lambda D,
\end{equation}
where $D$ is a diagonal matrix and the mixing proportion, $\lambda$, is determined by cross-validation. This estimator expresses the idea that in the absence of evidence for strong correlations between cells we assume they are independent (Fig.~1A). This approach is sometimes also referred to as ridge regression.

The second estimator we consider is shrinkage towards a low-rank matrix
\begin{equation}
\hat \Sigma_{\rm low-rank} = (1-\lambda) \hat \Sigma_0 + \lambda T,
\end{equation}
where $T = LL^\T + \Psi$ is a factor analysis model (REFS) consisting of a low-rank component $L$ with rank $d < n$ and a diagonal matrix $\Psi$. This estimator expresses the idea that the fluctuations in population activity are driven by a small number of unobserved sources that affect many cells (Fig.~1B).

The third estimator is based on sparsifying the inverse covariance (precision) matrix. It expresses the idea that, similar to synaptic connections, interactions between neurons are sparse (Fig.~1C). The estimator produces a low-dimensional approximation to the covariance that has many zeros in its inverse. This approximation problem is also known as covariance selection. The covariance matrix estimate is given by
\begin{equation}
\hat\Sigma = \hat S^{-1}
\quad\mbox{with}\quad
\hat S = \argmin\limits_{\|S\|_0 \le \rho} \loss{S^{-1},\hat\Sigma_0}   
\end{equation}
where $\mathcal L$ is a loss function (see below) and $\|S\|_0\le\rho$ signifies the constraint that $S$ has at most $\rho$ non-zero coefficients. While solving the problem in this form is computationally challenging, relaxing the $L_0$ norm to the $L_1$ norm converts it into one of convex optimization \cite{Donoho:2000}, which is computationally much more tractable. The resulting algorithm, which is known as graphical lasso \cite{Meinshausen:2006,Yuan:2007,Banerjee:2008,Friedman:2008}, produces the following estimator  
\begin{equation}
\hat\Sigma_{\rm sparse}^\mathcal{C} = \hat S^{-1}
\quad\mbox{with}\quad
\hat S = \argmin \loss{S^{-1},\hat\Sigma_0} + \lambda \|S\|_1,
\end{equation}
where $\|S\|_1$ is the $L_1$ norm of the precision matrix (see Methods for details) and $\lambda$ is again determined by cross-validation.

Finally, we consider a fourth estimator, which combines common inputs with sparse interactions (Fig.~1D). For this estimator the covariance is
\begin{equation}
\hat\Sigma_{\rm combined} = (S + LL^\T)^{-1},
\end{equation}
where, as above, $S$ is sparsified using $L_1$ norm and $L$ is a low-rank matrix with the rank determined by cross-validation.

\subsection*{Simulation using toy data}

To verify our approach and to illustrate the performance of our four regularized estimators and the sample covariance estimator, we constructed five model populations with different underlying structures. Each population contained 100~neurons. The first four populations matched the low-dimensional structure of our four estimators: independent (Fig.~2\,A), low-rank (Fig.~2\,B), sparse inverse (Fig.~2\,C) and low-rank combined with sparse inverse (Fig.~2\,D). \Acomment{Integrate schematics from Fig.~1 into this figure (make it the first row) for better correspondence} In addition, we considered a fifth population that had no low-dimensional structure (Fig.~2\,E).

To evaluate the performance of the different estimators, we computed the excess loss for all combinations of model populations and estimators, including the sample covariance. The first striking observation is that in all cases all four regularized estimators performed substantially better than the sample covariance (Fig.~2, fourth row). In particular, this is even true for the case where the population did not have any low-dimensional structure at all (Fig.~2\,E). While it may appear surprising at first sight, this counter-intuitive phenomenon is known as \emph{Stein's phenomenon} or \emph{Stein's paradox} \cite{Efron:1977}, named after its discoverer Charles Stein \cite{Stein:1956}. A common misconception about regularization is that its effect depends on accurate prior knowledge about the structure of the data. However, substantial improvement can be attained by shrinking the unbiased estimate toward an arbitrary target as long as the target is less variable than the unbiased estimator. The more accurate description of regularization is as of the optimal tradeoff between estimation and approximation error -- the so-called ``bias-variance tradeoff''.

Thus, if taken in isolation a regularized estimator improves the estimate we should not interpret this result to suggest that the estimator's target has the same low-dimensional structure as the data-generating process. However, when comparing multiple estimators against each other, the one whose target estimate most closely matches the true value with the smallest number of parameters will reduce the estimation error with the least increase in approximation error. Indeed, in all four toy examples with a low-dimensional structure, the estimator with the matching regularization target performed best (Fig.~2\,A--D, fourth row). Since the estimator combining low-rank and sparse interactions combines to low-dimensional targets and includes the simpler estimators as special cases, it performed almost as well even when the low-dimensional structure did not include either a low-rank component or sparse interactions (Fig.~2\,B,\,C). \Acomment{May want to include a smaller sample size as well since here the more reduced estimators (B, C) may actually outperform estimator D when the population structure matches their target and there is little data (say, 1.5x number of neurons)}

Since the above evaluations of excess loss require knowledge of ground truth, this analysis can be done only on toy data. However, an unbiased empirical estimate of the excess loss exists (see Methods, Eq.~\ref{eq:validationLoss} \Acomment{This may need a brief sentence about the intuition what's the difference, or a reference to the relevant methods section}).  To verify that our approach should also work in a realistic situation without access to ground truth, we also computed the empirical loss. Indeed, without access to ground truth, this analysis revealed a pattern of results that reproduced the results obtained with access to ground truth above (Fig.~2, last row). Because validation loss is computed by comparing estimates to noisy sample covariance matrices from smaller validation sets, empirical loss is a much noisier measurement than excess loss. In addition, it does not converge to zero with increasing sample size as excess loss does.

\subsection*{Covariance estimation on neural data}

We recorded the calcium activity of dense populations of neurons in the supragranular layers in primary visual cortex of anesthetized mice using fast random-access 3D scanning two-photon microscopy \cite{Stosiek:2003,Reddy:2005}. We presented numerous repetitions of full-field drifting gratings (Fig. 1A and 1B) to the eye contralateral to the imaged site. This technique allowed us to record from a large number (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4. We deconvolved somatic calcium signals using sparse nonnegative deconvolution \cite{Vogelstein:2010} (Fig.\;1C and 1D) and subtracted the average stimulus response to remove effects of the stimulus. From this residual response we computed the sample noise covariance matrix (Fig.\;1E).

In such highly localized populations both direct interactions between cells and common diffuse inputs are likely to contribute to the overall population variability. At the same time, most correlations are relatively small (Fig.~1\,E), suggesting that a simple shrinkage towards independence may provide a sufficiently well regularized estimate. By comparing the performance of our differently regularized estimators we can gain insights into which of these aspects are important in our data.

As expected, all regularized estimators outperformed the sample covariance estimator substantially (Fig.~4) \Acomment{I think we should show that}. Say something about how shrinkage, low-rank and sparse inverse relate \Acomment{Why do we do pairwise comparisons instead of just showing median (across sites) log loss relative to the combined estimator? I've seen many people do this and it provides an ordering. I feel like we should come up with a way of ordering them somehow, even if it's under certain assumptions that may not be entirely correct. I'm pretty sure if we don't do it the reviewers will bring it up anyway...}. Finally, the combined sparse and low-rank estimator dominated all others significantly (Fig.~4), showing that both hidden units and direct interactions are important in our data. The improvement from the low-rank estimator to the combined one is much larger than that from the sparse inverse to the combined one, suggesting that in this dataset direct interactions contribute more strongly to the correlation structure than hidden units do.

\subsection*{Relationship between functional covariance structure and circuit architecture}

\begin{itemize}
\item Linear and partial correlations versus spatial separation (lateral and vertical). Discuss whether to include or remove thresholded correlations.
\item Magnitude of common input versus spatial location in the volume (lateral and vertical separation from center).
\item Linear and partial correlations versus orientation preference
\item Distribution of sparsity over sites
\item Distribution of number of hidden units over sites
\item Distribution of eigenvalues for sample covariance versus combined estimator
\item etc. etc.

\end{itemize}
